Comparing Results with Mafessoni et al.

### 2019-07-09 LatexManuscript/legacy.tex
updates

\todo{All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.}

>>>>

All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

### 2019-07-09 PlottingScripts/ProcessedData/ScoopCheck.Rmd
updates

plt <- ggplot(data, 
              aes(x = End, 
                  y = Start, 
                  fill = Freq)) +
    geom_tile() +
    facet_grid(.~Mut) +
    labs(y = name) +
    theme_classic() +
    theme(axis.title.x = element_blank(),
          axis.line = element_blank(),
          axis.text = element_text(size = 20),
          axis.title.y = element_text(size = 20),
          strip.placement = "outside",
          strip.background = element_blank(),
          strip.text = element_text(size = 20),
          legend.text = element_text(size = 20),
          legend.title = element_text(size = 20),
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      size = 1),
          plot.margin = unit(c(0,0,0,1.5), "lines")) +
    scale_fill_gradient(limits = c(0, maxFreq),
                        low = 'white', 
                        high = 'royalblue1',
                        guide = guide_legend(title = 'Variant\ncount',
                                             direction = 'vertical'))
return(plt)
}
```

load data
```{r}
t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)
t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)

us <- fread('/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/Significant_Positions_and_Pvals.txt')

totalT<- unique(rbind(t4,t5,t6,t7,t8))
totalTus <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
totalTusX <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
totalTusX[is.na(totalTusX)] <- 1

```{r}
p_comp <- 
  ggplot(totalTusX, 
       aes(x=Adjusted_Log10p, 
           y= -log10(V5))) +
  geom_point(shape = 1) +
  labs(x = 'Q-association',
       y = 'LD-association')

p_comp
```

```{r}
mafessoni_snps <- 
  totalT[,c('V1','V3')]

>>>>

plt <- ggplot(data, 
              aes(x = End, 
                  y = Start, 
                  fill = Freq)) +
    geom_tile() +
    facet_grid(.~Mut) +
    labs(y = name) +
    theme_classic() +
    theme(axis.title.x = element_blank(),
          axis.line = element_blank(),
          axis.text = element_text(size = 20),
          axis.title.y = element_text(size = 20),
          strip.placement = "outside",
          strip.background = element_blank(),
          strip.text = element_text(size = 20),
          legend.text = element_text(size = 20),
          legend.title = element_text(size = 20),
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      size = 1),
          plot.margin = unit(c(0,0,0,1.5), "lines")) +
    scale_fill_gradient(limits = c(0, maxFreq),
                        low = 'white', 
                        high = 'black',
                        guide = guide_legend(title = 'Variant\ncount',
                                             direction = 'vertical'))
return(plt)
}
```

load data
```{r}
#load regression results
Regression <- readRDS('/Users/luke/Documents/Legacy_Misc/Misc/Regression.Rmd')
#load supplementary files from Mafessoni
#coding
t4 <- read.table('/Users/luke/Downloads/evy199_supp/TableS4.bed', comment.char = "@", header = T)
t4$type <- "rare_coding"
t5 <- read.table('/Users/luke/Downloads/evy199_supp/TableS5.bed', comment.char = "@", header = T)
t5$type <- "common_coding"
#intergenic
t6 <- read.table('/Users/luke/Downloads/evy199_supp/TableS6.bed', comment.char = "@", header = T)
t6$type <- "common_intergenic"
t7 <- read.table('/Users/luke/Downloads/evy199_supp/TableS7.bed', comment.char = "@", header = T)
t7$type <- "rare_intergenic"

#get total unique positions from most conservative tables
totalT<- 
  rbind(t4,t5,t6,t7) %>%
  group_by(X.chr, end) %>% 
  top_n(1) #keep unique positions only

merged_regression <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('X.chr','end'), all.x=T)
merged_regression[is.na(merged_regression)] <- 1

#remove positions that don't pass strict mask
merged_regression_masked <- 
  anti_join(merged_regression, 
            strictMask,
            by = c('Chr','Pos','rsID'))

#remove positions that DO pass strict mask
merged_regression_unmasked <- 
  anti_join(merged_regression, 
            merged_regression_masked)

nrow(merged_regression)

nrow(merged_regression_unmasked)

nrow(merged_regression_masked)

merged_regression_unmasked$strict_mask <- "Pass strict mask"
merged_regression_masked$strict_mask <- "Fail strict mask"

merged_regression <- rbind(merged_regression_masked, merged_regression_unmasked)
```

```{r}
#select rows of interest
#either present in Mafessoni OR significant in our data

to_plot <- merged_regression %>%
           filter(pvalue_glm_filt < 1 | Adjusted_Log10p > 2)

p_comp <- 
  ggplot(to_plot, 
         aes(x=Adjusted_Log10p, 
             y= -log10(pvalue_glm_filt), 
             color = strict_mask)) +
  geom_point(shape = 1,
             size = 1,
             alpha = 0.3) +
  labs(x = 'Q-association',
       y = 'LD-association') +
  scale_color_grey(start = 0.1, end = 0.6) +
  guides(color = F) +
  facet_wrap(.~strict_mask)+
  theme_classic() +
  theme(axis.title = element_text(size = 20),
        axis.text = element_text(size = 20))

p_comp
```
```{r}
#missing in Mafessoni
p <- to_plot %>%
  filter(strict_mask == 'Pass')

m_maf <- p %>%
           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)

#missing in ours
m_us <- p %>%
           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)

#present in both
both <- p %>%
           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)

print(paste("missing in Mafessoni (Pass): ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(p)*100) )
print(paste("missing in ours (Pass): ",nrow(m_us), "or : ",nrow(m_us)/nrow(p)*100))
print(paste("present in both (Pass): ",nrow(both), "or : ",nrow(both)/nrow(p)*100))
print(paste("total (Pass): ", nrow(p)))

#missing in Mafessoni
f <- to_plot %>%
  filter(strict_mask == 'Fail')

m_maf <- f %>%
           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)

#missing in ours
m_us <- f %>%
           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)

#present in both
both <- f %>%
           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)

print(paste("missing in Mafessoni (Fail): ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(f)*100) )
print(paste("missing in ours (Fail): ",nrow(m_us), "or : ",nrow(m_us)/nrow(f)*100))
print(paste("present in both (Fail): ",nrow(both), "or : ",nrow(both)/nrow(f)*100))
print(paste("total (Fail): ", nrow(f)))
```

```{r}
mafessoni_snps <- 
  totalT[,c('X.chr','end')]
